Connecting the Popularity Adjusted Block Model to the Generalized Random Dot Product Graph for Community Detection and Parameter Estimation

### 2021-06-23 summary.tex
updated document

For any edge probability matrix \(P\) for the PABM such that the rows
and columns are organized by community, the \(kl\)\textsuperscript{th}
block is an outer product of two vectors, i.e.,
\(P^{(kl)} = \lambda^{(kl)} (\lambda^{(lk)})^\top\). Therefore, given
\(P^{(kl)}\), \(\lambda^{(kl)}\) and \(\lambda^{(lk)}\) are solvable
up to multiplicative constant using singular value
decomposition. More specifically, let \(P^{(kl)} = \sigma^2 u v^\top\)
be the singular value decomposition of \(P^{(kl)}\).
\(u \in \mathbb{R}^{n_k}\) and \(v \in \mathbb{R}^{n_l}\) are vectors
and \(\sigma^2 > 0\) is a scalar. Then \(\lambda^{(kl)} = s_1 u\)
and \(\lambda^{(lk)} = s_2 v\) where $s_1 s_2 = \sigma^2$.
Given the adjacency matrix \(A\)
instead of edge probability matrix \(P\), we can simply use plug-in
estimators, which converge to the true parameters, again, up to a
multiplicative constant. Since these popularity vectors aren't identifiable,
due to the ambiguity of this constant,
we instead estimate each edge probability block
$P^{(kl)} = \lambda^{(kl)} (\lambda^{(lk)})^\top$
using the observed edge block $A^{(kl)}$ (Algorithm 3).
Alternatively, we can estimate the entire edge probability matrix as
$\hat{P} = \hat{X} I_{p,q} \hat{X}^\top$, where $\hat{X}$ is the ASE of $A$.

\begin{algorithm}[t]
  \DontPrintSemicolon
  \SetAlgoLined
  \caption{PABM block probability estimation.}
  \KwData{Adjacency block $A^{(kl)}$}
  \KwResult{Estimated edge probability block $\hat{P}^{(kl)}$}
  Compute $A^{(kl)} = U \Sigma V^\top$, the SVD of the $kl^{th}$ block.\;
  Assign $u^{(kl)}$ and $v^{(kl)}$ as the first columns of $U$ and $V$.
  Assign $(\sigma^{(kl)})^2 \leftarrow \Sigma_{11}$.\;
  Assign $\hat{\lambda}^{(kl)} \leftarrow \pm \sigma^{(kl)} u^{(kl)}$ and
  $\hat{\lambda}^{(lk)} \leftarrow \pm \sigma^{(kl)} v^{(kl)}$.\;
  Assign $\hat{P}^{(kl)} = \hat{\lambda}^{(kl)} (\hat{\lambda}^{(lk)})^\top$.
\end{algorithm}

\begin{theorem}
\label{theorem6}
Let $n_{kl} = n_k + n_l$.
Under regularity and sparsity assumptions, given fixed $K$,

\begin{equation} \label{eq:thm6}
\max_{k, l} \|\hat{P}_n^{(kl)} - P_n^{(kl)}\| = O_P \bigg(\frac{(\log n_{kl})}{\sqrt{n_{kl}}} \bigg)
\end{equation}
\end{theorem}

@@

We also define parameter estimation error as the RMSE
of the estimated edge probability blocks:

\[RMSE =
\sum_{k,l} \sqrt{\frac{1}{n_k n_l} \|P^{(kl)} - \hat{P}^{(kl)}\|_F^2}\]

@@

In each simulation, community labels \(z_1, ..., z_n\) were drawn from a
multinomial distribution with mixture parameters
\(\{\alpha_1, ..., \alpha_K\}\), then \(\{\lambda^{(kl)}\}_K\) according
to the drawn community labels, \(P\) was constructed using the drawn
\(\{\lambda^{(kl)}\}_K\), and \(A\) was drawn from \(P\) by
\(A_{ij} \stackrel{indep}{\sim} \Bernoulli(P_{ij})\). Each simulation has
a unique edge probability matrix \(P\).

@@

Given ground truth community labels, Algorithm 3 and the MLE-based
plug-in estimators \cite{307cbeb9b1be48299388437423d94bf1}
\cite{noroozi2019estimation} perform similarly, with root mean square
error decaying at rate approximately \(n^{-1/2}\).

@@

\emph{Case \(k = l\)}. \(P^{(kk)}_n\) and \(A^{(kk)}_n\) represent
within-community edge probabilities and edges for community \(k\).\\
By definition, \(P^{(kk)}_n = \lambda^{(kk)}_n (\lambda^{(kk)}_n)^\top\). This
implies that the singular value decomposition
\(P^{(kk)}_n = (\sigma^{(kk)}_n)^2 u^{(kk)}_n (u^{(kk)}_n)^\top\)
has one singular
value and one pair of singular vectors (\(P^{(kk)}_n\) is symmetric, so
the left and right singular vectors are identical). Then
\(\lambda^{(kk)}_n = \sigma^{(kk)}_n u^{(kk)}_n\).\\
Let \(\hat{U}^{(kk)}_n \hat{\Sigma}^{(kk)}_n (\hat{U}^{(kk)}_n)^\top\) be the
singular value decomposition of \(A^{(kk)}_n\), and let
\((\hat{\sigma}^{(kk)})^2 \hat{u}^{(kk)}_n (\hat{u}^{(kk)}_n)^\top\) be its
one-dimensional approximation. Define
\(\hat{\lambda}^{(kk)}_n = \hat{\sigma}^{(kk)}_n \hat{u}^{(kk)}_n\). Then
\(\hat{\lambda}^{(kk)}_n\) is the adjacency spectral embedding
approximation of \(\lambda^{(kk)}_n\).\\
Then by Theorem 5 of \citeauthor{rubindelanchy2017statistical}, the
adjacency spectral embedding \(\hat{\lambda}^{(kk)}_n\) approximates
\(\lambda^{(kk)}_n\) at rate \(\frac{(\log n_k)^c}{\sqrt{n_k}}\).

\emph{Case \(k \neq l\)}. \(P^{(kl)}_n\) and \(A^{(kl)}_n\) represent edge
probabilities and edges between communities \(k\) and \(l\). Note that
\(P^{(kl)} = (P^{(lk)})^\top\).\\
By definition, \(P^{(kl)} = \lambda^{(kl)} (\lambda^{(lk)})^\top\). As
in the \(k = l\) case, we note that the singular value decomposition
\(P^{(kl)} = \sigma_{kl}^2 u^{(kl)} (v^{(kl)})^\top\) is one-dimensional
and \(\lambda^{(kl)} = \sigma_{kl} u^{(kl)}\). (We can also note that
the SVD of \(P^{(lk)} = \sigma_{kl}^2 v^{(kl)} (u^{(kl)})^\top\), i.e.,
\(\sigma_{kl} = \sigma_{lk}\), \(u^{(kl)} = v^{(lk)}\), and
\(v^{(kl)} = u^{(lk)}\).)\\
Now consider the Hermitian dilation

@@

which is a symmetric \(n_{kl} \times n_{kl}\) matrix. It can
be shown that the spectral decomposition of \(M^{(kl)}\) is

@@

\[\hat{M}^{(kl)} = \begin{bmatrix} 0 & A^{(kl)} \\ A^{(lk)} & 0 \end{bmatrix}\]
Then \(\hat{M}^{(kl)}\) is an adjacency matrix drawn from \(M^{(kl)}\),
so its adjacency spectral embedding, given by

@@

where each \(\hat{\lambda}^{(kl)}\) is defined as in Algorithm 3,
approximates the latent positions of \(M^{(kl)}\) up to indefinite
orthogonal transformation with rate $\frac{(\log n_{kl})^c}{\sqrt{n_{kl}}}$
\cite{rubindelanchy2017statistical}.\\
In this case, the indefinite orthogonal transformation \(W_*\) in the
GRDPG result \cite{rubindelanchy2017statistical} is of the form
\(U^\top \hat{U}\). The eigenvalues of \(M\) are distinct since the
signature for this GRDPG is \((1, 1)\), and \(U^\top \hat{U}\) is block
diagonal, resulting in \(W_* \stackrel{a.s.}{\to} I\). Therefore, the
adjacency spectral embedding of \(\hat{M}^{(kl)}\) is a direct
estimation of the specific latent positions outlined for \(M^{(kl)}\),
up to sign flip.

>>>>

For any edge probability matrix \(P\) for the PABM such that the rows
and columns are organized by community, the \(kl\)\textsuperscript{th}
block is an outer product of two vectors, i.e.,
\(P^{(kl)} = \lambda^{(kl)} (\lambda^{(lk)})^\top\). Therefore, given
\(P^{(kl)}\), \(\lambda^{(kl)}\) and \(\lambda^{(lk)}\) are solvable
up to multiplicative constant using singular value
decomposition. More specifically, 
let \(P^{(kl)} = (\sigma^{(kl)})^2 u^{(kl)} (v^{(kl)})^\top\)
be the singular value decomposition of \(P^{(kl)}\).
\(u^{(kl)} \in \mathbb{R}^{n_k}\) and 
\(v^{(kl)} \in \mathbb{R}^{n_l}\) are vectors
and \(\sigma^{(kl)}\) is a scalar. Then \(\lambda^{(kl)} = s_1 u^{(kl)}\)
and \(\lambda^{(lk)} = s_2 v^{(kl)}\) 
for unidentifiable $s_1 s_2 = (\sigma^{(kl)})^2$.
Given the adjacency matrix \(A\)
instead of edge probability matrix \(P\), we can simply use plug-in
estimators by taking the SVD of each $A^{(kl)}$. 
Since each $\lambda^{(kl)}$ is not strictly identifiable,
we instead estimate each $\tilde{\lambda}^{(kl)} = \sigma^{(kl)} u^{(kl)}$.

\begin{algorithm}[t]
  \DontPrintSemicolon
  \SetAlgoLined
  \caption{PABM parameter estimation.}
  \KwData{Adjacency matrix $A$, community assignments $1, ..., K$}
  \KwResult{PABM parameter estimates $\{\hat{\lambda}^{(kl)}\}_K$.}
  Arrange the rows and columns of $A$ by community such that each 
  $A^{(kl)}$ block consists of estimated edge probabilities between 
  communities $k$ and $l$.\;
  \For {$k, l = 1, ..., K$, $k \leq l$} {
    Compute $A^{(kl)} = U \Sigma V^\top$, the SVD of the $kl^{th}$ 
    block.\;
    Assign $u^{(kl)}$ and $v^{(kl)}$ as the first columns of $U$ and $V$. 
    Assign $(\sigma^{(kl)})^2 \leftarrow \Sigma_{11}$.\;
    Assign $\hat{\lambda}^{(kl)} \leftarrow \pm \sigma^{(kl)} u^{(kl)}$ and 
    $\hat{\lambda}^{(lk)} \leftarrow \pm \sigma^{(kl)} v^{(kl)}$.
  }
\end{algorithm}

\begin{theorem}
\label{theorem6}
Under regularity and sparsity assumptions, given fixed $K$,

\begin{equation} \label{eq:thm6}
\max_{k, l \in \{1, ..., K\}} 
||\hat{\lambda}^{(kl)} - \lambda^{(kl)}|| = 
O_P \bigg(\frac{(\log n_k)^c}{\sqrt{n_k}} \bigg)
\end{equation}
\end{theorem}

@@

We also define two types of parameter estimation error. 
First, we estimate the popularity vectors directly and compute the RMSE. 
In this case, the "true" popularity vectors are derived from 
taking the SVD of each edge probability block $P^{(kl)}$ 
to avoid the unidentifiable multiplicative constants.

\[\text{RMSE}(\{\hat{\lambda}^{(kl)}\}_K, \{\lambda^{(kl)}\}_K) = 
\sqrt{\frac{1}{n} \sum_{k < l} \min_{s = \pm 1} 
\|s \hat{\lambda}^{(kl)} - \lambda^{(kl)} \|_2^2}\]

We can also avoid the unidentifiable multiplicative constant more directly 
by reconstructing each 
$\hat{P}^{(kl)} = \hat{\lambda}^{(kl)} (\hat{\lambda}^{(lk)})^\top$, 
which we use to define another parameter estimation error.

\[\text{RMSE}(\hat{P}, P) =
\sum_{k,l} \sqrt{\frac{1}{n_k n_l} \|P^{(kl)} - \hat{P}^{(kl)}\|_F^2}\]

@@

In each simulation, community labels \(z_1, ..., z_n\) were drawn from a
multinomial distribution with mixture parameters
\(\{\alpha_1, ..., \alpha_K\}\), then \(\{\lambda^{(kl)}\}_K\) according
to the drawn community labels, \(P\) was constructed using the drawn
\(\{\lambda^{(kl)}\}_K\), and \(A\) was drawn from \(P\) by
\(A_{ij} \indep \Bernoulli(P_{ij})\). Each simulation has
a unique edge probability matrix \(P\).

@@

Given ground truth community labels, Algorithm 3 and the MLE-based
plug-in estimators \cite{noroozi2019estimation, 307cbeb9b1be48299388437423d94bf1} 
perform similarly, with root mean square
error decaying at rate approximately \(n^{-1/2}\).

@@

\emph{Case \(k = l\)}. \(P^{(kk)}\) and \(A^{(kk)}\) represent
within-community edge probabilities and edges for community \(k\).\\
By definition, \(P^{(kk)} = \lambda^{(kk)} (\lambda^{(kk)})^\top\). This
implies that the singular value decomposition
\(P^{(kk)} = \sigma_{kk}^2 u^{(kk)} (u^{(kk)})^\top\) has one singular
value and one pair of singular vectors (\(P^{(kk)}\) is symmetric, so
the left and right singular vectors are identical). Then
\(\lambda^{(kk)} = \sigma_{kk} u^{(kk)}\).\\
Let \(\hat{U}^{(kk)} \hat{\Sigma}^{(kk)} (\hat{U}^{(kk)})^\top\) be the
singular value decomposition of \(A^{(kk)}\), and let
\(\hat{\sigma}_{kk}^2 \hat{u}^{(kk)} (\hat{u}^{(kk)})^\top\) be its
one-dimensional approximation. Define
\(\hat{\lambda}^{(kk)} = \hat{\sigma}_{kk} \hat{u}^{(kk)}\). Then
\(\hat{\lambda}^{(kk)}\) is the adjacency spectral embedding
approximation of \(\lambda^{(kk)}\).\\
Then by Theorem 5 of \citeauthor{rubindelanchy2017statistical}, the
adjacency spectral embedding \(\hat{\lambda}^{(kk)}\) approximates
\(\lambda^{(kk)}\) at rate \(\frac{(\log n_k)^c}{\sqrt{n_k}}\).

\emph{Case \(k \neq l\)}. \(P^{(kl)}\) and \(A^{(kl)}\) represent edge
probabilities and edges between communities \(k\) and \(l\). Note that
\(P^{(kl)} = (P^{(lk)})^\top\).\\
By definition, \(P^{(kl)} = \lambda^{(kl)} (\lambda^{(lk)})^\top\). As
in the \(k = l\) case, we note that the singular value decomposition
\(P^{(kl)} = \sigma_{kl}^2 u^{(kl)} (v^{(kl)})^\top\) is one-dimensional
and \(\lambda^{(kl)} = \sigma_{kl} u^{(kl)}\). (We can also note that
the SVD of \(P^{(lk)} = \sigma_{kl}^2 v^{(kl)} (u^{(kl)})^\top\), i.e.,
\(\sigma_{kl} = \sigma_{lk}\), \(u^{(kl)} = v^{(lk)}\), and
\(v^{(kl)} = u^{(lk)}\).)\\
Now consider the Hermitian dilation

@@

which is a symmetric \((n_k + n_l) \times (n_k + n_l)\) matrix. It can
be shown that the spectral decomposition of \(M^{(kl)}\) is

@@

\[\hat{M}^{(kl)} = \begin{bmatrix} 0 & A^{(kl)} \\ A^{(lk)} & 0 \end{bmatrix}\]
Then \(\hat{M}^{(kl)} = M^{(kl)} + E'\) where

\[E' = \begin{bmatrix} 0 & E \\ E^\top & 0 \end{bmatrix}\]

and \(E\) is the \(n_k \times n_l\) matrix of independent noise (to
generate the Bernoulli entries in \(A^{(kl)}\).\\
Then \(\hat{M}^{(kl)}\) is an adjacency matrix drawn from \(M^{(kl)}\),
so its adjacency spectral embedding, given by

@@

where each \(\hat{\lambda}^{(kl)}\) is defined as in Algorithm 3,
approximates the latent positions of \(M^{(kl)}\) up to indefinite
orthogonal transformation by the rate given in Theorem 5 of
\citeauthor{rubindelanchy2017statistical}.\\
In this case, the indefinite orthogonal transformation \(W_*\) in the
GRDPG result \cite{rubindelanchy2017statistical} is of the form
\(U^\top \hat{U}\). The eigenvalues of \(M\) are distinct since the
signature for this GRDPG is \((1, 1)\), and \(U^\top \hat{U}\) is block
diagonal, resulting in \(W_* \stackrel{a.s.}{\to} I\). Therefore, the
adjacency spectral embedding of \(\hat{M}^{(kl)}\) is a direct
estimation of the specific latent positions outlined for \(M^{(kl)}\),
up to sign flip.
